An Analysis of Health & Lifestyle Factors Associated with Sleep Quality

### 2024-04-16 products/manuscript/Manuscript.qmd
Manuscript Updates
Added figures, equations, an abstract, and updated some of the sections.

---
title: "An Analysis of Health & Lifestyle Factors Associated with Sleep Quality"
subtitle: ""
author: Kevin Kosewick and Emma Hardin-Parker
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../manuscript/dataanalysis-references.bib

---

*ANYTHING BELOW CODE LINE 172 IS A DRAFT AND IS NOT TO BE CONSIDERED A PART OF THE FINAL MANUSCRIPT*

@@

  *WILL NEED COLLAB FOR PART 5*

@@

  *NEED TO INSERT *

@@

  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8345503/ --> blood pressure and sleep
  https://www.apa.org/news/press/releases/stress/2013/sleep --> sleep duration & stress on sleep quality 
  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10503965/#:~:text=Regular%20physical%20activity%20can%20lead,managing%20sleep%20disorders%20like%20insomnia. --> physical activity and sleep quality

@@

The dataset originally contained 373 observations for 13 variables covering a broad spectrum health and lifestyle variables associated with sleep quality. The curators of the dataset did not provide information on how or where this data was collected, so we solely relied on the codebook for variable definitions. There were no missing values or erroneous variables in the dataset that we needed to remove. The "Blood.Pressure" variable was transformed to reflect the systolic and diastolic blood pressure measurements from each subject. We then chose to add an additional variable, labeled 'cat_bp' to reflect the categorical blood pressure status of a subject based on the American Heart Association guidelines *INCLUDE REFERENCE HERE (can be found in eda.qmd file)* . Additionally we created a variable, 'StepsGroup' to determine the categorical activity level of subjects in the dataset based on these guidelines provided by 10,000 steps *[10,000 steps.org](https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500).* Finally, we created an additional categorical variable, 'PhysicalActivityGroup' based on 30 minute increments, to categorically represent the level of physical activity reported by a subject. The outcome of interest is noted as "Quality.of.Sleep" that we will reference as Sleep Quality or Quality of Sleep throughout this report.

*Unsure if we need this to be here? Or if it will even show up given the FALSE arguments we added*
```{r Package Loading, warning=FALSE, message=FALSE}
#load in the data and check it out
library(here)
library(tidyverse)
library(dplyr)
library(skimr)
```

## Variables included in Analyses *Include Questions / Hypotheses Here*

@@

@@

Random Forest Models and Cross-Validation were considered to be the "best" fit for our outcome of interest, Quality of Sleep. The data were not split into train/test subsets as the data contains less than 400 observations and many of the values were unique in comparison to the rest of the data. Prior to constructing this model, a collinearity plot was constructed to determine any presence of collinearity in our data. *INCLUDE PLOT HERE* Stress and Sleep Quality had an absolute value of 0.9 on the correlation scale. Sleep Duration and Stress were also seen to be strongly correlated. Stress was still included in our models; however, we found it important to note the strong linear relationship between Stress and Sleep Quality.

A variable importance plot (VIP) was constructed to determine which variables had the strongest impact in our Random Forest Model. *INCLUDE PLOT HERE* Sleep Duration, Stress Level, and Age had the highest levels of importance in our model. Physical Activity Level was excluded entirely, which was to be expected given our exploratory data analysis and simple linear regression models.

@@

Root Mean Squared Error (RMSE) is a common metric used to evaluate regression models. The RMSE is formally defined as the square root of the mean square of all error and is defined by the following formula: *Include LaTeX equation of RMSE?* . It is important that while RMSE is scale dependent, common practice notes that low RMSE values indicate stronger model performance.

@@

The 'Quality of Sleep' variable has been selected as our outcome of interest. It is important to note that this variable is a subjective measure of a participant's self-reported sleep quality. The following plot shows that most subjects reported a sleep quality score of 8 (out of 10). *REFERENCE SLEEP_DIS PLOT FROM EDA.QMD FILE*

@@

A variable importance plot (VIP) was constructed to determine which variables had the strongest impact in our Random Forest Model. *INCLUDE PLOT HERE* Sleep Duration, Stress Level, and Age had the highest levels of importance in our model. Physical Activity Level was excluded entirely, which was to be expected given our exploratory data analysis and simple linear regression models.

@@

# Discussion

@@

  https://link.springer.com/article/10.1007/s11325-015-1193-z --> obesity and sleep quality (affects men but not women in China)

  https://www.scielosp.org/article/gs/2017.v31n1/18-22/en/?uid=680b54f39b# --> effect of age and gender on sleep quality (women worse than men, sleep quality deteriorates with age in women but is a bit all over the place for men)

  https://academic.oup.com/sleep/article/33/2/149/2454438#77919376 --> sleep quality by occupation/industry (managerial positions had the worst score)

*ANYTHING BELOW THIS COMMENT IS FROM PRIOR TO THE OVERHAUL AND IS STILL TO BE IGNORED / MOVED TO ANOTHER SECTION OF THE PROJECT*

## Questions/Hypotheses to be addressed

We intend to explore the influence of each variable on sleep quality and identify which ones seem to have the most impact (both positive and negative). We also wish to explore if certain variables affect male sleep quality more than females. If time allows, we'd also like to see if certain ages seem to be affected by certain variables more often than others. We can focus on every predictor in the dataset since there are only 13.

We can likely use GLMs and t-tests/ANOVA tests to see the significance that variables may have in different combinations on an individual's sleep quality. Heat maps and various plots will also help with exploratory data analysis. We aren't exceptionally experienced with statistical analyses, but these are what we can think of at the moment.

## Data import

We'll load in the cleaned data here. Please reference the eda.qmd and Supplementary-Material.qmd files for full exploratory results and the cleaning process.

```{r Loading the Processed Data}
#load the cleaned data
sleepdata<- readRDS(here("data","processed-data", "sleepdataprocessed.rds"))

```

## Statistical analysis

We'll begin to look at some of the relationships between our variables now. Sleep Quality is our outcome of interest for all of these. We'll start with Physical Activity level.

```{r}
############################
#### Fitting a model using Quality of Sleep as the outcome and Physical Activity Level as a predictor

lmfit_quality_activity <- lm(Quality.of.Sleep ~ Physical.Activity.Level, sleepdata)

# Placing results from lmfit_quality_activty  into a data frame with the tidy function
lmtable_quality_activity <- broom::tidy(lmfit_quality_activity)

# Viewing the results from the first model fit 
print(lmtable_quality_activity)

# Saving the lmfit_quality_activity results table  
lmtable_quality_activity = here("results", "tables", "lmfit1table.rds")
saveRDS(lmtable_quality_activity, file = lmtable_quality_activity)
```

The intercept value indicates that Quality of Sleep will be 6.66 if physical activity is at 0. This has a relatively low standard error and high significance.The coefficient for our variable indicates that as Physical activity level increases by one unit, Quality of Sleep will increase by 0.0109. Our t-statistic is indicated as significant by the p-value but is much lower than the intercept's.This means that Physical Activity Level does have a measurable impact on Quality of Sleep, but it's relatively small.

Next, we'll look at Sleep Duration as a predictor.

```{r}
############################
#### Fitting a model using Quality of Sleep as the outcome and Sleep Duration as a predictor

lmfit_quality_duration <- lm(Quality.of.Sleep ~ Sleep.Duration, sleepdata)

# Placing results from lmfit_quality_duration into a data frame with the tidy function
lmtable_quality_duration <- broom::tidy(lmfit_quality_duration)

#Viewing the results from the second model fit 
print(lmtable_quality_duration)

# Saving the lmfit_quality_duration results table  
lmtable_quality_duration = here("results", "tables", "lmfit2table.rds")
saveRDS(lmtable_quality_duration, file = lmtable_quality_duration)

```

For our sleep duration model, we can see that our intercept's negative value indicates that Quality of Sleep would be very poor if individuals got no sleep. This is rational and is supported by a strong p-value and a decent t-statistic, although the standard error is a bit high. For our variable coefficient, we can see that as sleep duration increases by one unit, Quality of sleep also increases by about 1. This has a very strong p-value and t-statistic, indicating a strong relationship between this predictor and our outcome of interest.

Next, we'll look at Stress Level as our predictor.

```{r}
############################
####Fitting a model using Sleep Quality as the outcome and Stress Level as a predictor.

lmfit_quality_stress <- lm(Quality.of.Sleep~ Stress.Level, sleepdata)

#Placing results from lmfit_quality_stress into a data frame with the tidy function
lmtable_quality_stress <- broom::tidy(lmfit_quality_stress)

#Viewing the results from the sixth model fit 
print(lmtable_quality_stress)

# Saving the lmfit_quality_stress results table  
lmtable_quality_stress = here("results", "tables", "lmfit6table.rds")
saveRDS(lmtable_quality_stress, file = lmtable_quality_stress)
```

According to our model, Stress has a strongly defined relationship with sleep quality. We can see that a value of 0 on the stress score leads to a 10 unit increase in quality of sleep. The Stress Level coefficient indicates that as stress increases by one unit, sleep quality decreases by about 0.6 of a unit. The t-statistics and p-values for the values in this model are extremely strong, indicating a well-defined relationship between stress and sleep quality.

Finally, we'll look at Age as our predictor.

```{r}
############################
####Fitting a model using Sleep Quality as the outcome and Age as a predictor.

lmfit_quality_age <- lm(Quality.of.Sleep~ Age, sleepdata)

#Placing results from lmfit_quality_gender_age into a data frame with the tidy function
lmtable_quality_age <- broom::tidy(lmfit_quality_age)

#Viewing the results from the seventh model fit 
print(lmtable_quality_age)

# Saving the lmfit_quality_gender_age results table  
lmtable_quality_age = here("results", "tables", "lmfit8table.rds")
saveRDS(lmtable_quality_age, file = lmtable_quality_age)

```

Our model shows a strong relationship between age and quality of sleep. As Age increases, sleep quality seems to improve. Each year increase in age is predicted to have a 0.065 increase in sleep quality score. The intercept here is also interesting as an age of "0" is predicted to have a sleep score of 4.57, which is relatively poor. The t-statistics and p-values for these are both strong.

## Exploratory/Descriptive analysis

Here are the most interesting/distinct plots from the EDA:

Physical Activity plot:

We can clearly see that Physical Activity level has a positive correlation with Quality of Sleep.
```{r Physical Activity Plot, warning=FALSE, message=FALSE}
#load plot package
library(ggplot2)

# Box plot for Quality of Sleep by Physical Activity Level Group
ggplot(sleepdata, aes(x = PhysicalActivityGroup, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Physical Activity Level Group", x = "Physical Activity Level Group", y = "Quality of Sleep")
```

Gender, Age, and Occupation:

We can see some clear differences in self-reported Quality of Sleep between these different groups already.
Females tend to report an average of 1 higher according to the boxplot.
The different age groups are similar, but the oldest (50-60) reports the highest quality of sleep by far.
Our occupations are a bit all over the place, but Sales Representatives seem to have the worst average scores by far.
```{r Gender, Age, and Occupation Plot}
# Boxplot for Quality of Sleep by Gender
ggplot(sleepdata, aes(x = Gender, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Gender", x = "Gender", y = "Quality of Sleep")

# Create bins for Age in order to create a clean boxplot.
sleepdata$AgeGroup <- cut(sleepdata$Age, breaks = c(20, 30, 40, 50, 60), labels = c("20-30", "30-40", "40-50", "50-60"), include.lowest = TRUE)

# Calculate mean Quality of Sleep for each Age Group to plot with means for cleaner visuals
sleepdata_summary <- sleepdata %>%
  group_by(AgeGroup) %>%
  summarise(MeanQuality = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for Mean Quality of Sleep by Age Group with colorblind-friendly palette
ggplot(sleepdata_summary, aes(x = AgeGroup, y = MeanQuality, fill = AgeGroup)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean Quality of Sleep by Age Group", x = "Age Group", y = "Mean Quality of Sleep")

# Calculate mean Quality of Sleep for each Occupation in order to create a bar plot.
sleepdata_summary <- sleepdata %>%
  group_by(Occupation) %>%
  summarise(MeanQuality = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for Mean Quality of Sleep by Occupation with manually specified colors
ggplot(sleepdata_summary, aes(x = Occupation, y = MeanQuality, fill = Occupation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = rainbow(length(unique(sleepdata_summary$Occupation)))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mean Quality of Sleep by Occupation", x = "Occupation", y = "Mean Quality of Sleep")
```

BMI:
We can clearly see that Normal weights seem to have better sleep on average than the obese and overweight categories. However, there are very few in the obese category, so conclusions with that group may not be as supported as others.

```{r BMI plot}
# Box plot for Quality of Sleep by BMI Category
ggplot(sleepdata, aes(x = BMI.Category, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by BMI Category", x = "BMI Category", y = "Quality of Sleep")
```

Blood Pressure:

The boxplots look a bit strange for the Elevated and Normal levels, but we can see that it's likely due to a shortage of data for these categories. We can see, however, that Quality of Sleep does seem to be negatively correlated with increasing stages of hypertension.

```{r Blood Pressure Plot}
# Bar plot for cat_bp
ggplot(sleepdata, aes(x = cat_bp)) +
  geom_bar() +
  labs(title = "Bar Plot of cat_bp", x = "cat_bp", y = "Count")

# Box plot for Quality of Sleep by cat_bp
ggplot(sleepdata, aes(x = cat_bp, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by cat_bp", x = "cat_bp", y = "Quality of Sleep")

```

And Stress Level:
Our scatterplot shows a clear negative correlation between Stress Level and Quality of Sleep. However, this is a subjective scale, so data should be taken with a grain of salt for now.

```{r Stress Level Plot}
# Scatterplot for Quality of Sleep by Stress Level
ggplot(sleepdata, aes(x = Stress.Level, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Quality of Sleep by Stress Level", x = "Stress Level (scale: 1-10)", y = "Quality of Sleep")
```

The other plots were either self explanatory (e.g. Sleep Duration meant higher sleep quality) or had less defined trends (e.g. heart rate and daily steps).

>>>>

---
title: "An Analysis of Health & Lifestyle Factors Associated with Sleep Quality"
subtitle: ""
author: Kevin Kosewick and Emma Hardin-Parker
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../manuscript/dataanalysis-references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

For our project, we obtained a dataset from Kaggle.com containing 373
observations of 13 variables from a study on different factors that may
affect sleep quality. Variables of interest from the study included
physical activity level, age, sex, occupation, stress, and several more.
In RStudio (version 4.3), we used a combination of packages to create
models suitable for answering our primary question: which variables are
most important for determining an individual's sleep quality and what
effect do they have? After conducting an exploratory data analysis (EDA)
we hypothesized that stress, sleep duration, and occupation would have
the largest impact on sleep quality. We believed that stress and sleep
quality would be negatively correlated, sleep duration and quality would
be positively correlated, and that jobs with longer or more variable
hours would negatively affect sleep quality.

After creating linear models for each variable's impact on sleep quality
and fitting a Random Forest model with Cross-Validation to test variable
importance, we determined that sleep duration, stress level, and age
were the most important factors. Our hypothesis for sleep duration and
stress was supported; however, we did not correctly predict that age
would be one of the most important predictors for sleep quality. Our
hypotheses regarding the effects of sleep duration, stress, and
occupation were all supported as well. Overall, we determined that an
individual's duration of sleep, stress level, and age are all key
factors that can impact sleep quality.

*Feel free to edit this Emma; I tried to include everything needed*

@@

@@

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8345503/ --\> blood
pressure and sleep
https://www.apa.org/news/press/releases/stress/2013/sleep --\> sleep
duration & stress on sleep quality
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10503965/#:\~:text=Regular%20physical%20activity%20can%20lead,managing%20sleep%20disorders%20like%20insomnia.
--\> physical activity and sleep quality

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum health and lifestyle variables associated with
sleep quality. The curators of the dataset did not provide information
on how or where this data was collected, so we solely relied on the
codebook for variable definitions. There were no missing values or
erroneous variables in the dataset that we needed to remove. The
"Blood.Pressure" variable was transformed to reflect the systolic and
diastolic blood pressure measurements from each subject. We then chose
to add an additional variable, labeled 'cat_bp' to reflect the
categorical blood pressure status of a subject based on the [*American
Heart Association
guidelines*](https://www.heart.org/en/health-topics/high-blood-pressure).
Additionally we created a variable, 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by 10,000 steps [*10,000
steps.org*](https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500)*.*
Finally, we created an additional categorical variable,
'PhysicalActivityGroup' based on 30 minute increments, to categorically
represent the level of physical activity reported by a subject. The
outcome of interest is noted as "Quality.of.Sleep" that we will
reference as Sleep Quality or Quality of Sleep throughout this report.

## Variables included in Analyses

@@

The objective of our analysis was to determine which of these variables
are most important for determining an individual's sleep quality and
what specific effect they have. After conducting an (EDA) we
hypothesized that stress, sleep duration, and occupation would have the
largest impact on sleep quality. We believed that stress and sleep
quality would be negatively correlated, sleep duration and quality would
be positively correlated, and that jobs with longer or more variable
hours would negatively affect sleep quality.

@@

Random Forest Models and Cross-Validation were considered to be the
"best" fit for our outcome of interest, Quality of Sleep. The data were
not split into train/test subsets as the data contains less than 400
observations and many of the values were unique in comparison to the
rest of the data. Prior to constructing this model, a collinearity plot
was constructed to determine any presence of collinearity in our data.
![Collinearity between all variables.](images/collinearity.png)

Stress and Sleep Quality had an absolute value of 0.9 on the correlation
scale. Sleep Duration and Stress were also seen to be strongly
correlated. Stress was still included in our models due to its
well-studied impact on sleep; however, we found it important to note the
strong linear relationship between Stress and Sleep Quality.

A variable importance plot (VIP) was constructed to determine which
variables had the strongest impact in our Random Forest Model. ![The
ranking of each variable\'s importance in determining sleep quality
according to our Random Forest
model.](images/variableimportance.png)Sleep Duration, Stress Level, and
Age had the highest levels of importance in our model. Physical Activity
Level was excluded entirely, which was to be expected given our
exploratory data analysis and simple linear regression models.

@@

Root Mean Squared Error (RMSE) is a common metric used to evaluate
regression models. The RMSE is formally defined as the square root of
the mean square of all error and is defined by the following formula:
$$\sum_{i=1}^{D}(x_i-y_i)^2$$

It is important to consider that while RMSE is scale dependent, common
practice notes that low RMSE values indicate stronger model performance.

R-squared (R²) is a statistical measure used in regression analyses. It
quantifies the proportion of the variance in a dependent variable that
can be explained by an independent variable within a regression model
and is calculated with the following formula:

$$
[ R^2 = 1 - \frac{{\text{{Unexplained Variation}}}}{{\text{{Total Variation}}}} ]
$$

We used both RMSE and R² to evaluate the performance of our model
compared to a "null model". Null models generate predictions for an
outcome of interest without using any predictors. Instead, they simply
average the values of the outcome of interest to make "predictions".

![The results generated from our RF model. RMSE increases when more
predictors are added to the model. The R-squared value also decreases as
more predictors are added to the model.](images/RFtuningresults.png)

Figure (*name/order figures after we finalize draft*) shows that as we
include more predictors in our model, RMSE increases slightly while the
R-squared value decreases. This indicates that some variables in our
dataset are more important for predicting sleep quality than the others.
Based off of our variable importance graph, we can assume that stress
level, sleep duration, and age are the factors that carry the most
weight when predicting sleep quality.

We went on to evaluate how well our RF model predicted values by
creating an observed vs. predicted plot.

![A comparison of the actual values for sleep quality in our dataset
(observed) compared to the values predicted by our fitted RF
model.](images/observed_vs_predicted.png)

While the predicted values don't align perfectly with the red "ideal
value" line, they fall very close overall, indicating that our model can
sufficiently predict sleep quality using unseen data.

@@

The 'Quality of Sleep' variable has been selected as our outcome of
interest. It is important to note that this variable is a subjective
measure of a participant's self-reported sleep quality. The following
plot shows that most subjects reported a sleep quality score of 8 (out
of 10).

![Distribution of Sleep Quality Scores in the
dataset.](images/unnamed-chunk-17-1.png)

@@

*To Emma: did you want this text in both places you placed it?* A
variable importance plot (VIP) was constructed to determine which
variables had the strongest impact in our Random Forest Model. *INCLUDE
PLOT HERE* Sleep Duration, Stress Level, and Age had the highest levels
of importance in our model. Physical Activity Level was excluded
entirely, which was to be expected given our exploratory data analysis
and simple linear regression models.

@@

# Discussion (5)

@@

https://link.springer.com/article/10.1007/s11325-015-1193-z --\> obesity
and sleep quality (affects men but not women in China)

https://www.scielosp.org/article/gs/2017.v31n1/18-22/en/?uid=680b54f39b#
--\> effect of age and gender on sleep quality (women worse than men,
sleep quality deteriorates with age in women but is a bit all over the
place for men)

https://academic.oup.com/sleep/article/33/2/149/2454438#77919376 --\>
sleep quality by occupation/industry (managerial positions had the worst
score)
